Geometrization of characters of tori over non-Archimedean local fields

### 2013-07-12 shanghai.tex
Working on switching to TikZ

% !TEX encoding = UTF-8 Unicode
\documentclass[11pt]{amsart}
\title[Geometration of characters of tori]{Geometrization of characters of tori over non-Archimedean local fields}
\usepackage[british]{babel}
\usepackage{datetime}
\date{\today}
\author{Clifton Cunningham}
\address{University of Calgary}
\email{cunning@math.ucalgary.ca}
\author{David Roe}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
\email{roed.math@gmail.com}
\usepackage[utf8]{inputenc}
\renewcommand{\baselinestretch}{1.2}
\usepackage[notcite,color]{showkeys}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[shortalphabetic]{amsrefs}
\usepackage[alphabetic]{amsrefs}
\renewcommand\MR{\relax}
\usepackage{xypic}
\usepackage{textcomp}
\usepackage{mathrsfs}
\usepackage{yfonts}
\newcommand{\mathswab}[1]{\operatorname{\textswab{#1}}}
\usepackage[T1]{fontenc}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc}
\usepackage{manfnt}

@@

\tikzset{every picture/.style={>=stealth}}

@@

  Let us review how line bundles on $G^0$ determine character sheaves
  on $G^0$.  Up to isomorphism, line bundles on $G^0$ may be
  apprehended in terms of pairs $(\pi,\psi)$,
  \David{Aren't there line bundles which are not character sheaves,
    for which $B$ is not an algebraic group?}
  where $\pi : B\to G^0$ is a finite etale morphism of algebraic
  groups and a character $\psi : A\to \EEx$, where
  $\Gal(\bFq/\Fq)$ acts trivially on $A \ceq \ker {\bar \pi}$, as
  summarized by the following diagram.
  \begin{equation}\label{E}
    \xymatrix{
      \EEx & 1\ar[r] & \ar@/_1pc/[ll]_{\psi} A  \ar[r] & B \ar[r]^{\pi} & G^0 \ar[r] & 1\\
      }
  \end{equation}
  (In fact, we may always take $B = G^0$ and $\pi = \Lang_{G^0}$, the
  Lang isogeny for $G^0$.)  Indeed, $\cs{E}$ is precisely the
  component of $\pi_! (\EE)_{B}$ on which $A$ acts through $\psi$.  To
  see that $\cs{E}$ is a line bundle on $G^0$, start by pulling back
  $\pi: B \to G^0$ along $m_0 : \GzxGz \to G^0$.
  \begin{equation}
    \xymatrix{
      \EEx & 1\ar[r] & \ar@/_1pc/[ll]_{\psi} A  \ar[r] & B \ar[r]^{\pi} & G^0 \ar[r] & 1\\
      & 1\ar[r] & \ar@{.>}@/^1pc/[llu] \ar[u] \AxA  \ar[r]
      & \ar[u]^{m_{B}} \BxB \ar[r]^{\pi\tight{1}{\times}{1} \pi}  & \ar[u]^{m_0} \GzxGz \ar[r] & 1\\
      }
  \end{equation}
  As pictured above, the kernel of this pull-back is $\AxA$;
  indeed, $\AxA \to \BxB$ is the pull-back of $A\to B$ along
  $m_{B}$. Use this Cartesian square to define a character (the dotted
  arrow, above) of $\AxA$; that is, define $\AxA \to \EEx$
  by the composition of $\AxA \to A$ and $\psi : A\to \EEx$.
  Note that this character is $\psi \otimes \psi : \AxA \to \EEx$.
  \Clifton{I haven't explained how maps between diagrams define sheaf
    homomorphisms between local systems.}
  Since the lower sequence is precisely the sequence defining
  $\ExE$, this shows that, by construction, $\cs{E}$
  comes equipped with a canonical isomorphism $m_0^*\cs{E} \iso \ExE$.
  Now the base change $\gcs{E}$ comes equipped with isomorphisms
  $F^0 : \Frob_{G^0}^* \gcs{E} \to \gcs{E}$ and
  $\mu^0 : \bm_0^* \gcs{E} \to \bExE$
  satisfying conditions \ref{CS.1} and \ref{CS.2}. Then $(\gcs{E},\mu^0,F^0)$
  is a character sheaf of $G^0$.

  The next step is to show that $\pi: B \to G^0$ extends to a finite
  etale morphism of group schemes $f: H\to G$, smooth and locally of
  finite type over $\Fq$, in such a way that $H^0 = B$, $f^0 =\pi$ and
  $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism of component
  groups.
  \begin{equation} \label{extension-diagram}
    \xymatrix{
      & A \ar[d] \ar@{=}[r] & A \ar[d] &  \\
      & B \ar[r] \ar[d]_{\pi} & H \ar[r] \ar[d]_{f} & \pi_0(H) \ar[d]_{\pi_0(f)}^{\iso} \\
      & G^0 \ar[r] & G \ar[r] & \pi_0(G).
    }
  \end{equation}
  where all rows and columns are exact and all maps are defined over
  $\Fq$.  We will do so by passing back and forth between group
  schemes over $\Fq$ and their $\bFq$-points.  Extensions of
  $G^0(\bFq)$ by $A(\bFq)$, such as $A(\bFq) \to H^0(\bFq) \to G^0(\bFq)$,
  correspond to classes in $\Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq))$.
  Similarly, extensions of $G(\bFq)$ by $A(\bFq)$ correspond to
  classes in $\Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq))$.  The map
  $G^0(\bFq) \to G(\bFq)$ induces a map
  $\Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq))$,
  fitting into the long exact sequence derived from applying
  $\Hom(\mbox{---}, A(\bFq))$ to $G^0(\bFq) \to G(\bFq) \to \pi_0(G)(\bFq)$:
  $$\Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq)) \to \Ext^2_{\ZZ[\Gamma]}(\pi_0(G)(\bFq), A(\bFq)).$$
  We now show that $\Ext^2_{\ZZ[\Gamma]}(\pi_0(G)(\bFq), A(\bFq)) = 0$.

  It suffices to show that $\pi_0(G)(\bFq)$ has a two-term resolution
  by injective $\ZZ[\Gamma]$-modules.  Since the action of $\Gamma$ on
  $\pi_0(G)(\bFq)$ is continuous, it factor through a finite quotient
  $\Gamma_m = \Gamma / H_m$.  Any product of Dedekind domains has
  cohomological dimension $1$ \cite[?]{?}, so $\pi_0(G)(\bFq)$ has a
  two term resolution by injective $\ZZ[\Gamma_m]$-modules.  Let $I$
  be one of these modules: we need to show that $I$ is injective as a
  $\ZZ[\Gamma]$-module.  Consider the universal diagram for injectivity:
  \[
  \xymatrix{
    & B \\
    I & A \ar@{>->}[u] \ar[l]_{a}
    }
  \]

  where $A$ and $B$ are arbitrary $\ZZ[\Gamma]$-modules.  Since $H_m$ acts trivially on $I$,
  $I_{H_m}A := \langle (\gamma-1)a : \gamma \in H_m, a \in A \rangle \subseteq \ker(a).$
  Since $B / I_{H_m}B \cong B \otimes_A A / I_{H_m}A$, we get
  \[
  \xymatrix{
    & B/I_{H_m}B \ar@{..>}[dl] & B \ar@{->>}[l] \\
    I & A/I_{H_m}A \ar@{>->}[u] \ar[l] & A \ar@{>->}[u] \ar@{->>}[l]
    }
  \]
  The dotted arrow exists since $I$ is injective as a
  $\ZZ[\Gamma_m]$-module, and the outer diagram shows that $I$ is
  injective as a $\ZZ[\Gamma]$-module.

@@

@@

  \begin{tikzpicture}[auto,style={rectangle, scale=1.2}]
  \node (12) [anchor=east] at (4,2) {$\Aut(\bf^{x+y})$};
  \node (22) [anchor=west] at (4.4,2) {$\bH^{x+y}$};
  \node (32) [anchor=east] at (11,2) {$\bG^{x+y}$};
  \node (01) at (0,1) {$\EEx$};
  \node (11) [anchor=east] at (4,1) {$\Aut(\bf^{x+y})$};
  \node (21) [anchor=west] at (4.4,1) {$\bH^{x+y} \times_{\bG^x\times\bG^y} \bG^x\times\bG^y$};
  \node (31) [anchor=east] at (11,1) {$\bG^x\times \bG^y$};
  \node (10) [anchor=east] at (4,0) {$\Aut(\bf^x)\times \Aut(\bf^y)$};
  \node (20) [anchor=west] at (4.4,0) {$\bG^x\times \bG^y$};
  \node (30) [anchor=east] at (11,0) {$\bG^x\times \bG^y$};
  \draw [->] (4.3,1.94) arc (330:30:0.12);
  \draw [->, densely dotted] (4.3,0.94) arc (330:30:0.12);
  \draw [->] (4.3,-0.06) arc (330:30:0.12);
  \draw [->] (12) to node[swap] {\footnotesize{$\psi_{x+y}$}} (01);
  \draw [->, densely dotted] (11) -- (01);
  \draw [->] (10) to node {\footnotesize{$\psi_x \otimes \psi_y$}} (01);
  \draw [->, densely dotted] ($(11.south) + (0,-0.5)$) -- (11);
  \draw [double equal sign distance] (11) -- (12);
  \draw [->] (22) to node {\footnotesize{$\bf^{x+y}$}} (32);
  \draw [->] ($(22.south) + (0,-0.5)$) -- (22);
  \draw [->] (21) -- (31);
  \draw [->] ($(32.south) + (0,-0.5)$) to node[swap] {\footnotesize{$\bm_{x,y}$}} (32);
  \draw [->] ($(22.south) + (0,-1.5)$) -- ($(22.south) + (0,-1)$);
  \draw [->] (20) to node {\footnotesize{$\bf^x \times \bf^y$}} (30);
  \draw [->] ($(32.south) + (0,-1.5)$) -- ($(32.south) + (0,-1)$);
  \end{tikzpicture}

>>>>

% !TEX encoding = UTF-8 Unicode
\documentclass[11pt]{amsart}
\title[Geometration of characters of tori]{Geometrization of characters of tori over non-Archimedean local fields}
\usepackage[british]{babel}
\usepackage{datetime}
\date{\today}
\author{Clifton Cunningham}
\address{University of Calgary}
\email{cunning@math.ucalgary.ca}
\author{David Roe}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
\email{roed.math@gmail.com}
\usepackage[utf8]{inputenc}
\renewcommand{\baselinestretch}{1.2}
\usepackage[notcite,color]{showkeys}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[shortalphabetic]{amsrefs}
\usepackage[alphabetic]{amsrefs}
\renewcommand\MR{\relax}
\usepackage{xypic}
\usepackage{textcomp}
\usepackage{mathrsfs}
\usepackage{yfonts}
\newcommand{\mathswab}[1]{\operatorname{\textswab{#1}}}
\usepackage[T1]{fontenc}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{manfnt}

@@

\tikzset{every picture/.style={>=stealth},label/.style={font=\footnotesize}}

@@

  Let us review how line bundles on $G^0$ determine character sheaves
  on $G^0$.  Up to isomorphism, line bundles on $G^0$ may be
  apprehended in terms of pairs $(\pi,\psi)$,
  \David{Aren't there line bundles which are not character sheaves,
    for which $B$ is not an algebraic group?}
  where $\pi : B\to G^0$ is a finite etale morphism of algebraic
  groups and a character $\psi : A\to \EEx$, where
  $\Gal(\bFq/\Fq)$ acts trivially on $A \ceq \ker {\bar \pi}$, as
  summarized by the following diagram.
  \begin{equation}\label{E}
    \xymatrix{
      \EEx & 1\ar[r] & \ar@/_1pc/[ll]_{\psi} A  \ar[r] & B \ar[r]^{\pi} & G^0 \ar[r] & 1\\
      }
  \end{equation}
  (In fact, we may always take $B = G^0$ and $\pi = \Lang_{G^0}$, the
  Lang isogeny for $G^0$.)  Indeed, $\cs{E}$ is precisely the
  component of $\pi_! (\EE)_{B}$ on which $A$ acts through $\psi$.  To
  see that $\cs{E}$ is a line bundle on $G^0$, start by pulling back
  $\pi: B \to G^0$ along $m_0 : \GzxGz \to G^0$.

  \begin{center}
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=2em] {
  \EEx & 1 & A & B & G^0 & 1 \\
  & 1 & \AxA & \BxB & \GzxGz & 1\\ };
  %\foreach \x in {2,3,4,5} \let \x1={\x+1} 
  %\draw [->] (m-1-\x) -- (m-1-\x1);
  \end{tikzpicture}
  \end{center}

  \begin{equation}
    \xymatrix{
      \EEx & 1\ar[r] & \ar@/_1pc/[ll]_{\psi} A  \ar[r] & B \ar[r]^{\pi} & G^0 \ar[r] & 1\\
      & 1\ar[r] & \ar@{.>}@/^1pc/[llu] \ar[u] \AxA  \ar[r]
      & \ar[u]^{m_{B}} \BxB \ar[r]^{\pi\tight{1}{\times}{1} \pi}  & \ar[u]^{m_0} \GzxGz \ar[r] & 1\\
      }
  \end{equation}
  As pictured above, the kernel of this pull-back is $\AxA$;
  indeed, $\AxA \to \BxB$ is the pull-back of $A\to B$ along
  $m_{B}$. Use this Cartesian square to define a character (the dotted
  arrow, above) of $\AxA$; that is, define $\AxA \to \EEx$
  by the composition of $\AxA \to A$ and $\psi : A\to \EEx$.
  Note that this character is $\psi \otimes \psi : \AxA \to \EEx$.
  \Clifton{I haven't explained how maps between diagrams define sheaf
    homomorphisms between local systems.}
  Since the lower sequence is precisely the sequence defining
  $\ExE$, this shows that, by construction, $\cs{E}$
  comes equipped with a canonical isomorphism $m_0^*\cs{E} \iso \ExE$.
  Now the base change $\gcs{E}$ comes equipped with isomorphisms
  $F^0 : \Frob_{G^0}^* \gcs{E} \to \gcs{E}$ and
  $\mu^0 : \bm_0^* \gcs{E} \to \bExE$
  satisfying conditions \ref{CS.1} and \ref{CS.2}. Then $(\gcs{E},\mu^0,F^0)$
  is a character sheaf of $G^0$.

  The next step is to show that $\pi: B \to G^0$ extends to a finite
  etale morphism of group schemes $f: H\to G$, smooth and locally of
  finite type over $\Fq$, in such a way that $H^0 = B$, $f^0 =\pi$ and
  $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism of component
  groups.

  \begin{center}
  \begin{tikzpicture}[auto]
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=2em] {
  A & A & \\
  B & H & \pi_0(H) \\
  G^0 & G & \pi_0(G) \\};
  \draw [double equal sign distance] (m-1-1) -- (m-1-2);
  \draw [->] (m-1-1) -- (m-2-1);
  \draw [->] (m-2-1) -- (m-2-2);
  \draw [->] (m-1-2) -- (m-2-2);
  \draw [->] (m-2-1) to node[label,swap] {$\pi$} (m-3-1);
  \draw [->] (m-3-1) -- (m-3-2);
  \draw [->] (m-2-2) to node[label,swap] {$f$} (m-3-2);
  \draw [->] (m-2-2) -- (m-2-3);
  \draw [->] (m-3-2) -- (m-3-3);
  \draw [->] (m-2-3) to node[below,rotate=90] {$\sim$} node[label,swap] {$\pi_0(f)$} (m-3-3);
  \end{tikzpicture}
  \end{center}

  \begin{equation} \label{extension-diagram}
    \xymatrix{
      & A \ar[d] \ar@{=}[r] & A \ar[d] &  \\
      & B \ar[r] \ar[d]_{\pi} & H \ar[r] \ar[d]_{f} & \pi_0(H) \ar[d]_{\pi_0(f)}^{\iso} \\
      & G^0 \ar[r] & G \ar[r] & \pi_0(G).
    }
  \end{equation}
  where all rows and columns are exact and all maps are defined over
  $\Fq$.  We will do so by passing back and forth between group
  schemes over $\Fq$ and their $\bFq$-points.  Extensions of
  $G^0(\bFq)$ by $A(\bFq)$, such as $A(\bFq) \to H^0(\bFq) \to G^0(\bFq)$,
  correspond to classes in $\Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq))$.
  Similarly, extensions of $G(\bFq)$ by $A(\bFq)$ correspond to
  classes in $\Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq))$.  The map
  $G^0(\bFq) \to G(\bFq)$ induces a map
  $\Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq))$,
  fitting into the long exact sequence derived from applying
  $\Hom(\mbox{---}, A(\bFq))$ to $G^0(\bFq) \to G(\bFq) \to \pi_0(G)(\bFq)$:
  $$\Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq)) \to \Ext^2_{\ZZ[\Gamma]}(\pi_0(G)(\bFq), A(\bFq)).$$
  We now show that $\Ext^2_{\ZZ[\Gamma]}(\pi_0(G)(\bFq), A(\bFq)) = 0$.

  It suffices to show that $\pi_0(G)(\bFq)$ has a two-term resolution
  by injective $\ZZ[\Gamma]$-modules.  Since the action of $\Gamma$ on
  $\pi_0(G)(\bFq)$ is continuous, it factor through a finite quotient
  $\Gamma_m = \Gamma / H_m$.  Any product of Dedekind domains has
  cohomological dimension $1$ \cite[?]{?}, so $\pi_0(G)(\bFq)$ has a
  two term resolution by injective $\ZZ[\Gamma_m]$-modules.  Let $I$
  be one of these modules: we need to show that $I$ is injective as a
  $\ZZ[\Gamma]$-module.  Consider the universal diagram for injectivity:

  \begin{center}
  \begin{tikzpicture}[auto]
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=2em] {
  & B \\
  I & A \\};
  \draw [right hook->] (m-2-2) -- (m-1-2);
  \draw [->] (m-2-2) to node[swap,label] {$a$} (m-2-1);
  \end{tikzpicture}
  \end{center}

  \[
  \xymatrix{
    & B \\
    I & A \ar@{>->}[u] \ar[l]_{a}
    }
  \]

  where $A$ and $B$ are arbitrary $\ZZ[\Gamma]$-modules.  Since $H_m$ acts trivially on $I$,
  $I_{H_m}A := \langle (\gamma-1)a : \gamma \in H_m, a \in A \rangle \subseteq \ker(a).$
  Since $B / I_{H_m}B \cong B \otimes_A A / I_{H_m}A$, we get

  \begin{center}
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=2em] {
  & B/I_{H_m}B & B \\
  I & A/I_{H_m}A & A \\};
  \draw [->,dashed] (m-1-2) -- (m-2-1);
  \draw [right hook->] (m-2-2) -- (m-1-2);
  \draw [->] (m-2-2) -- (m-2-1);
  \draw [->>] (m-2-3) -- (m-2-2);
  \draw [right hook->] (m-2-3) -- (m-1-3);
  \draw [->>] (m-1-3) -- (m-1-2);
  \end{tikzpicture}
  \end{center}

  \[
  \xymatrix{
    & B/I_{H_m}B \ar@{..>}[dl] & B \ar@{->>}[l] \\
    I & A/I_{H_m}A \ar@{>->}[u] \ar[l] & A \ar@{>->}[u] \ar@{->>}[l]
    }
  \]
  The dotted arrow exists since $I$ is injective as a
  $\ZZ[\Gamma_m]$-module, and the outer diagram shows that $I$ is
  injective as a $\ZZ[\Gamma]$-module.

@@

  \begin{tikzpicture}[auto,style={rectangle}]
  \node (0) at (0,0) {$\EEx$};
  \node (1) [anchor=east] at (7,0) {$\Aut(\bf^x) \times \Aut(\bf^y)$};
  \node (2) [anchor=west] at (7.4,0) {$\bH^x \times \bH^y$};
  \node (3) at (12,0) {$\bG^x \times \bG^y$};
  \draw [->] (1) to node[swap, label] {$\psi_x \otimes \psi_y$} (0);
  \draw [->] (7.3,-0.06) arc (330:40:0.12) -- (7.3,0.07);
  \draw [->] (2) to node[label] {$\bf^x \tight{0}{\times}{0} \bf^y$} (3);
  \end{tikzpicture}

@@

  \begin{tikzpicture}[auto,style={rectangle},scale=1.2]
  \node (12) [anchor=east] at (4,2) {$\Aut(\bf^{x+y})$};
  \node (22) [anchor=west] at (4.4,2) {$\bH^{x+y}$};
  \node (32) [anchor=east] at (11,2) {$\bG^{x+y}$};
  \node (01) at (0,1) {$\EEx$};
  \node (11) [anchor=east] at (4,1) {$\Aut(\bf^{x+y})$};
  \node (21) [anchor=west] at (4.4,1) {$\bH^{x+y} \times_{\bG^x\times\bG^y} \bG^x\times\bG^y$};
  \node (31) [anchor=east] at (11,1) {$\bG^x\times \bG^y$};
  \node (10) [anchor=east] at (4,0) {$\Aut(\bf^x)\times \Aut(\bf^y)$};
  \node (20) [anchor=west] at (4.4,0) {$\bG^x\times \bG^y$};
  \node (30) [anchor=east] at (11,0) {$\bG^x\times \bG^y$};
  \draw [->] (4.3,1.94) arc (330:40:0.12) -- (4.3,2.07);
  \draw [->, densely dotted] (4.3,0.94) arc (330:40:0.12) -- (4.3,1.07);
  \draw [->] (4.3,-0.06) arc (330:40:0.12) -- (4.3,0.07);
  \draw [->] (12) to node[swap,label] {$\psi_{x+y}$} (01);
  \draw [->, densely dotted] (11) -- (01);
  \draw [->] (10) to node[label] {$\psi_x \otimes \psi_y$} (01);
  \draw [->, densely dotted] ($(11.south) + (0,-0.5)$) -- (11);
  \draw [double equal sign distance] (11) -- (12);
  \draw [->] (22) to node[label] {$\bf^{x+y}$} (32);
  \draw [->] ($(22.south) + (0,-0.5)$) -- (22);
  \draw [->] (21) -- (31);
  \draw [->] ($(32.225) + (0,-0.5)$) to node[swap] {\footnotesize{$\bm_{x,y}$}} (32.225);
  \draw [->] ($(22.south) + (0,-1.5)$) -- ($(22.south) + (0,-1)$);
  \draw [->] (20) to node[label] {$\bf^x \times \bf^y$} (30);
  \draw [->] ($(32.225) + (0,-1.5)$) -- ($(32.225) + (0,-1)$);
  \end{tikzpicture}
